Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-13 paper/ResponseToReviews.tex

{\color{red}
Gini index is something we could definitely calculate. Not clear what we could compare it to, but we can at least calculate it and report it.
We then calculate Gini index under some simple model where everything is due to decreased transcription when missing segments.
Any such model is probably pretty bogus and oversimplified, so we'd want to do this in a way where everything ends up in a heavily caveated supplement that doesn't bog down text.
}

2. While the silent tagging method used to address co-infection is clever and appreciated, the issue is not fully settled. The dismissal of co-infection as a factor influencing cell-to-cell variation in lines 127-9 is based on too few cells to draw any conclusions, and thus needs to be tempered. Also, there is likely to be a significant amount of cryptic co-infection with identical barcode viruses (expected to be similar to that of mixed barcodes, ~10\%) that could influence measured heterogeneity. These points should be made in the text.

{\color{black}
These are very good points.
We have added the following text that clearly draws out the fact that our approach will only identify half the co-infections:
\begin{quote}
\textsl{
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure 4D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
}
\end{quote}

The reviewers are correct that our statement about the role of co-infection in influencing cell-to-cell variation was too strong.
We have clarified this statement.
We have also added a figure supplement (Figure 4-Figure supplement 4) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
The new text reads:
\begin{quote}
\textsl{
Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 4).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 4).
}
\end{quote}
}

@@

5. The determination of the minimum required influenza fraction (Figure 4) is based on sound logic; however, we are concerned that the observed results do not fully align with this model. Specifically, while the 10hr experiment looks reasonable, the distributions of wildtype/synonymous mixed cells in other experiments do not appear to show the same trend (most notable for 6hr and 8hr samples, where almost no mixed cells are observed at low fractions). In that regard, using the 10hr dataset to estimate thresholds for all of the other samples does not seem appropriate.

{\color{black}
This is a good point. 
We now use the procedure in Figure 4C to call thresholds \emph{separately} for each sample as the reviewers suggest.
The sample-specific calling is shown in Figure 4-Figure supplement 3.
These new thresholds lead to a modest increase in the number of cells called as virally infected for the earlier timepoint samples.
These newly called cells therefore slightly shift the quantitative values in all subsequent analyses, although the changes are very small and all results remain unchanged.

The reason that the earlier timepoints have less mixed barcodes is because these earlier samples have much less total viral mRNA -- and when their is less viral mRNA in total, there is also less available to be acquired from lysed cells.
}

@@

-Figure 3 is not effective in supporting the observations made in the text (that most cells have little viral RNA, and that there's a wide variability in the amount of viral RNA). Separate distributions for the viral RNA would be more telling. Log scale would be more appropriate for showing counts per cell.

{\color{red} Needs to be explaiend}

-Figure 4C: the threshold line appears to be a bit to the right of the global maximum - going through the description, it was not clear why that should be the case.

{\color{black} 
This is a good observation. 
We were determining the threshold as the maximum of a loess fit to the data. 
Previously, the loess fitting parameters caused the fit line to overshoot, putting the threshold a bit to the right.
We have adjusted span parameter in the loess fitting, which fixes the overshoot.
The revised Figure 4C shows the loess fit in orange as well as the threshold in dotted green -- and the threshold now goes right through the maximum.
}

### 2018-01-13 paper/paper.tex
fixed threshold for infected cells

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a cutoff for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
We fit a curve to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
See the Methods for details.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
}
\figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
The number of viral barcodes called for each sample and gene segment. 
Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
For calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
}{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.
}{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}

@@

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and may miss some true low-level infections.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure~\ref{fig:viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

@@

We also quantified the fraction of infected cells that completely failed to express a given gene.
We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C).
The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C). 
If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 40-50\% of infected cells express mRNAs from all eight genes.
This estimate of the frequency at which infected cells express mRNAs from all eight gene segments is slightly higher than previous estimates of 13\%~\citep{Brooke:2013kb} and 20\%~\citep{Dou:2017cp}.
At least one difference is that \citet{Brooke:2013kb} stained for proteins whereas we examined the expression of mRNAs -- it is likely that some cells contain mutated viral genes that fail to produce stable protein even when mRNA is expressed.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_manualedits.pdf}}
\caption{\label{fig:fluexpr}
Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
{\bf (A)} 
The fraction of viral mRNA from each viral gene for each cell. 
{\bf (B)}
Box plots showing the distribution of the fraction of viral mRNA per cell from each viral gene.
The black lines at the notches are the medians, and the tops and bottoms of boxes indicate the first and third quartiles.
Whiskers extend to the highest or lowest data point observed within 1.5x the interquartile range, outliers shown as circles.
Notches extend 1.58x the interquartile range divided by the square root of the number of observations. 
{\bf (C)}, {\bf (D)} 
The same plots, but only including cells for which we observed at least one molecule of each viral gene.
}
\figdata{The raw data for all cells are in \texttt{p\_flu\_expr\_all.csv}.}
\figdata{The raw data for fully infected cells are in \texttt{p\_flu\_expr\_fullyinfected.csv}.}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge_manualarrow.pdf}}
\caption{\label{fig:tsne}
A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
Each point is a single cell, and each panel shows an identical layout but colors the cells according to a different property.
{\bf (A), (B)}
Cells colored by the fraction of their mRNA derived from virus.
{\bf (C)}
Cells colored by the experimental sample.
{\bf (D), (E)}
Cells colored by the number of detected transcripts from type I and III interferons (IFN).
Only one cell has detectable interferon expression (in orange, indicated with arrow).
{\bf (E)}
Cells colored by the expression of the interferon-stimulated gene IFIT1.
{\bf (F)}
Cells colored by whether they express the viral NS gene.
The one interferon-positive cell is lacking NS, but so are many interferon-negative cells.
}
\end{figure}

@@

\subsection{Some host genes co-vary with viral gene expression.}
We examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap_manualedit.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

@@

%\nocite{*} % This command displays all refs in the bib file
\bibliography{references}

>>>>

\begin{figure}[t!]
\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a threshold for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
This plot illustrates the process for the 10-hour sample, see Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds} for similar plots for other samples.
See the Methods for details.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
}
\figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
The number of viral barcodes called for each sample and gene segment. 
Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
For calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
}{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
\figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
Cell lysis can lead cells to the spurious association of small amounts of extraneous mRNA with individual cells.
We wanted to avoid classifying as infected cells that had simply acquired such lysis-derived viral mRNA.
The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
As is shown in Figure~\ref{fig:viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
We called the threshold at this point of maximum purity (dotted green line).
For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as Figure~\ref{fig:viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (Figure~\ref{fig:cells}B).
}{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}
\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.
}{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}

@@

We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and may miss some true low-level infections.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure~\ref{fig:viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

@@

We also quantified the fraction of infected cells that completely failed to express a given gene.
We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C and Figure~\ref{fig:fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C and Figure~\ref{fig:fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}). 
If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 35-50\% of infected cells express mRNAs from all eight genes.
This estimate of the frequency at which infected cells express mRNAs from all eight gene segments is slightly higher than previous estimates of 13\%~\citep{Brooke:2013kb} and 20\%~\citep{Dou:2017cp}.
At least one difference is that \citet{Brooke:2013kb} stained for proteins whereas we examined the expression of mRNAs -- it is likely that some cells contain mutated viral genes that fail to produce stable protein even when mRNA is expressed.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr.pdf}}
\caption{\label{fig:fluexpr}
\jdbcomment{add expression wedges}
Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
{\bf (A)} 
The fraction of viral mRNA from each viral gene for each cell. 
{\bf (B)}
Box plots showing the distribution of the fraction of viral mRNA per cell from each viral gene.
The black lines at the notches are the medians, and the tops and bottoms of boxes indicate the first and third quartiles.
Whiskers extend to the highest or lowest data point observed within 1.5x the interquartile range, outliers shown as circles.
Notches extend 1.58x the interquartile range divided by the square root of the number of observations. 
{\bf (C)}, {\bf (D)} 
The same plots, but only including cells for which we observed at least one molecule of each viral gene.
}
\figdata{The raw data for all cells are in \texttt{p\_flu\_expr\_all.csv}.}
\figdata{The raw data for fully infected cells are in \texttt{p\_flu\_expr\_fullyinfected.csv}.}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge.pdf}}
\caption{\label{fig:tsne}
\jdbcomment{add arrow}
A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
Each point is a single cell, and each panel shows an identical layout but colors the cells according to a different property.
{\bf (A), (B)}
Cells colored by the fraction of their mRNA derived from virus.
{\bf (C)}
Cells colored by the experimental sample.
{\bf (D), (E)}
Cells colored by the number of detected transcripts from type I and III interferons (IFN).
Only one cell has detectable interferon expression (in orange, indicated with arrow).
{\bf (E)}
Cells colored by the expression of the interferon-stimulated gene IFIT1.
{\bf (F)}
Cells colored by whether they express the viral NS gene.
The one interferon-positive cell is lacking NS, but so are many interferon-negative cells.
}
\end{figure}

@@

\subsection{Some host genes co-vary with viral gene expression.}
We examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
We identified 43 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
\jdbcomment{add gray arrow wedge}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

@@

\bibliography{references}
